Implementing Reactive Systems Directly\\with I/O Automata in ioa++

### 2012-02-23 papers/paper2/design.tex
Removing material.

\begin{figure}
\center
\includegraphics[width=\columnwidth]{architecture}
\caption{Framework architecture.}
\label{framework_architecture}
\end{figure}

@@

The scheduler is responsible for both selecting and executing local actions.
These two activities are decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.

The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
Input actions are executed with corresponding output actions, according to the current set of bindings.
The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.

Since concurrent access to the dispatcher is allowed, the dispatcher implementation must be thread-safe.
Our implementation of the dispatcher uses a two-level locking scheme to enforce the concurrent execution constraints.
Any action that modifies the set of automata and bindings must first acquire a write lock to prevent concurrent access.
All other actions acquire a read lock to implement single-writer/multiple-reader semantics.
The second phase of executing a non-modifying action involves acquiring locks for the set of automata implied by an action.
The locks are acquired in a fixed order to prevent deadlock~\cite{havender1968avoiding}.
Once the locks are acquired, the precondition is evaluated, the local action and all bound input actions are executed, and the locks are released.

The controller encodes a \emph{policy} that selects the next local action.
For the framework to be a valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
The run-time system can be initialized with different controllers to achieve different performance objectives.
To leverage the power of different operating system mechanisms without exposing users to their complexities directly, the ioa++ framework is designed so that all native concurrency and synchronization mechanisms (e.g., threads and mutexes) are encapsulated within the dispatcher and controller.
For example, a controller can be written to use multiple threads of execution to exploit fully the concurrency inherently available in a system of automata,
while still providing concurrency semantics in a manner consistent with the I/O automata component model.

\paragraph*{Explicit dynamic scheduling}
The I/O automata model assumes that the scheduler knows about every local action in the system.
There are two approaches for achieving this in practice.
In the first approach, automata declare \emph{statically} all of their actions to the scheduler.
The controller then has the responsibility of cycling through the actions according to some fair policy.
Since the scheduler has no knowledge about which actions are \emph{enabled} or \emph{disabled} (have a precondition that is true or false respectively), this approach reduces to naively trying each action.

A more focused approach, which is the approach used in our framework implementation, is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
%The act of declaring an action to the scheduler is called \emph{scheduling}.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare actions to the scheduler (i.e., to \emph{schedule} them) after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

In this approach, scheduling disabled actions is unnecessary.
In order for a scheduled disabled action $a$ to become enabled, the automaton associated with the action must change state by executing another enabled action $b$ sometime after $a$ is scheduled.
Consequently, the scheduling of the disabled action $a$ can be deferred until the execution of some other action.
Notice that refraining from scheduling disabled actions does not affect the fairness of the scheduler since they cannot cause a state change when selected.
This result leads to a guarded programming idiom where the precondition of each action is checked before it is added to the schedule.

However, an automaton must still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
A simple technique for accomplishing this is to schedule all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
However, this must be done carefully, since not considering an enabled action would violate the fairness of the controller.
% since it will never select an action that may cause a state change.
% Not scheduling an (enabled) action is a common mistake.

\paragraph*{Binding predicates and scheduling}
%% This assumption is not implicit because we state that the preconditions are independent.
The preceding discussion on scheduling assumed that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind action changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after each successful bind or unbind.

\paragraph*{Actions of no-longer-existing automata}
Every action in the set of actions considered by the controller is scheduled by its associated automaton.
Since automata can be destroyed, an action in the set owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier assuming that the set of automata in the system turn over less frequently overall than the set of actions that have been scheduled.

\paragraph*{Delayed actions}
Scheduling activities for a future time is a useful feature for many systems because it allows the system to sleep between activities. For example, the framework should support the development of automata that act as timers and alarms.
Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph*{File descriptor events}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects may adversely affect the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is asynchronously ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
The ioa++ framework uses the Reactor pattern since (1) it is simpler and (2) many operating systems have non-blocking I/O and some kind of synchronous I/O (de)multiplexing, whereas asynchronous I/O is less widespread.

@@

\paragraph*{Basic and multi-threaded schedulers}
The ioa++ framework distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.
Scheduling is idempotent to prevent unbounded growth of the queue, and both scheduler implementations use the Reactor pattern to implement delayed actions and file descriptor events.

@@

Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with an indication of either success or failure.
Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
Subjects allow an automaton to have multiple requests outstanding at once.

Automata use a simple protocol for creating and destroying child automata.
Each child automaton is represented by a unique subject which is in one or two of five states: CreateSend, CreateRecv, CreateDone, DestroySend, and DestroyRecv.
Subjects start in the CreateSend state which indicates that the request for this subject is waiting to be sent to the system automaton.
Once the request has been sent, the subject transitions to the CreateRecv state and waits for a response from the system automaton.
Once the system automaton sends the response, the subject transitions to the CreateDone state.
When the user automaton requests that the child automaton be destroyed, the subject is either (1) added to the DestroySend state if the subject is in CreateRecv or CreateDone or (2) removed from all states if the subject is in CreateSend.
A subject appearing in both CreateDone and DestroySend transitions to the DestroyRecv state when the automaton makes the request to destroy it.
The system automaton can process system actions in any order.
This protocol ensures that an automaton does not request that a child automaton be destroyed until it receives confirmation that it was created.
The subject is forgotten when the system automaton responds with the result of the destroy action.

The protocol for binding is similar but addresses the possibility that a binding can be dissolved at any point due to the asynchronous destruction of automata.
We define analogous states for the binding state machine: BindSend, BindRecv, BindDone, UnbindSend, and UnbindRecv.
A binding subject can be in either BindRecv, BindDone, UnbindSend, or UnbindRecv when it receives a result from the system automaton indicating that the binding was dissolved.

\paragraph*{Automaton and Binding managers}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple constellations but becomes unwieldy when creating more complicated ones.
Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.

\ifjournal
\paragraph*{Generators}
A \emph{generator} is a single-use factory for creating an instance of an automaton.
Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.
\fi

>>>>

\begin{figure}
\center
\includegraphics[width=.6\columnwidth]{architecture}
\caption{Framework architecture.}
\label{framework_architecture}
\end{figure}

@@

The scheduler is responsible for both selecting and executing local actions.
These two activities are decomposed into a \emph{dispatcher} that executes actions and a \emph{controller} that selects the next action.
The dispatcher enforces the execution and concurrency constraints of Section~\ref{component_model}.
Input actions are executed with corresponding output actions, according to the current set of bindings.
The dispatcher allows actions to be executed concurrently by (1)~requiring that the sets of automata implied by each action are disjoint and (2)~serializing all updates to the set of automata and bindings.
The controller encodes a \emph{policy} that selects the next local action.
For the framework to be a valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
The run-time system can be initialized with different controllers to achieve different performance objectives.
The ioa++ framework distribution provides both a basic scheduler that is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics, and a multi-threaded scheduler with a configurable number of threads that we use in the performance evaluation described in Section~\ref{evaluation}.

\paragraph*{Explicit dynamic scheduling.}
In our framework implementation, automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
%The act of declaring an action to the scheduler is called \emph{scheduling}.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare actions to the scheduler (i.e., to \emph{schedule} them) after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.
Scheduling is idempotent to prevent unbounded growth of scheduler data structures.

\paragraph*{Binding predicates and scheduling.}
%% This assumption is not implicit because we state that the preconditions are independent.
The preceding discussion on scheduling assumed that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind action changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after each successful bind or unbind.

\paragraph*{Delayed actions.}
Scheduling activities for a future time is a useful feature for many systems because it allows the system to sleep between activities. For example, the framework should support the development of automata that act as timers and alarms.
Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph*{File descriptor events.}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects may adversely affect the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is asynchronously ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
The ioa++ framework uses the Reactor pattern since (1) it is simpler and (2) many operating systems have non-blocking I/O and some kind of synchronous I/O (de)multiplexing, whereas asynchronous I/O is less widespread.

@@

@@

Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with an indication of either success or failure.
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple constellations but becomes unwieldy when creating more complicated ones.
Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.

\ifjournal
\paragraph*{Generators.}
A \emph{generator} is a single-use factory for creating an instance of an automaton.
Generators are used to produce new automata as part of the create system action and are visible as arguments in the various layers of the framework.
\fi

### 2012-02-23 papers/paper2/paper.tex
Removing material.

\documentclass[10pt, conference, compsocconf]{IEEEtran}

@@

\title{Implementing Reactive Systems Directly\\with I/O Automata in ioa++}
\author{
  \IEEEauthorblockN{Justin R. Wilson and Christopher D. Gill}
  \IEEEauthorblockA{Department of Computer Science and Engineering\\
    Washington University in St. Louis\\
    Campus Box 1045\\
    One Brookings Drive\\
    St. Louis, MO 63130\\
    \{wilsonj,cdgill\}@cse.wustl.edu}
  \and
  \IEEEauthorblockN{Gruia-Catalin Roman}
  \IEEEauthorblockA{School of Engineering\\
    University of New Mexico\\
    Centennial Engineering Center, Suite 3071\\
    1 University of New Mexico, MSC01 1140\\
    Albuquerque, NM 87131-0001\\
    gcroman@unm.edu}
}

@@

\begin{IEEEkeywords}
concurrency; I/O automata; components
\end{IEEEkeywords}

@@

\bibliographystyle{IEEEtran}
\bibliography{bibliography}{}

>>>>

\documentclass{llncs}

@@

\title{Implementing Reactive Systems Directly\\with I/O Automata in ioa++}
\author{Justin R. Wilson\inst{1} \and Christopher D. Gill\inst{1} \and Gruia-Catalin Roman\inst{2}}

\institute{
  Dept. of Computer Science and Engineering, 
  Washington University, St. Louis, MO
\and
  School of Engineering,
  University of New Mexico, Albuquerque, NM
  \{wilsonj,cdgill\}@cse.wustl.edu gcroman@unm.edu
}

@@

\keywords{concurrency, I/O automata, components}

@@

\bibliographystyle{splncs03}
\bibliography{bibliography}{}

### 2012-02-23 papers/paper2/programming_model.tex
Removing material.

\paragraph*{Declaring and initializing state variables}
\fi
The state variables of an automaton are expressed as member variables of its corresponding class or class template.
\ifjournal
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status.
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
\fi
Member variables are declared private, which is in keeping with the notion that the state of each automaton is independent, and are initialized using a constructor.
The constructor can also bootstrap the scheduler using the ioa::schedule function discussed below.
\ifjournal
For the asynch\_lcr\_automaton, the flags are both set initially false and the UID is initialized and added to the send queue.
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the ioa::schedule function discussed below.
\fi

\paragraph*{Declaring/defining actions}
Actions are member variables that model the \emph{action} 
\ifjournal
concept.
\else
concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
\fi
Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
These traits are used to guide the execution of the methods by the dispatcher and to supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

@@

\paragraph*{Scheduling}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common idiom is for each scheduling function in an automaton to call a common member function that schedules all enabled local actions to avoid replicating scheduling logic.
The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{scheduling}.

@@

\paragraph*{Dynamic composition}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in Section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class also contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
\fi

\paragraph*{Binding predicates}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of other actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of output actions to ensure that the value generated by the output action will be received if produced.
\ifjournal
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi

@@

\paragraph*{System action events}
Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in Section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.

\paragraph*{Error reporting}
The event semantics of I/O automata provide a natural mechanism for reporting errors.
An automaton that encounters an error can use an output action to signal the error.
The automaton can continue processing if the error is transient or stop if the error is permanent.
In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.

\paragraph*{Handling errors by resetting}
An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
Automaton designers should strive to write automata that are capable of being reset as this is the simplest error handling strategy.

\paragraph*{Handling errors by replacement}
The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
Recall that system actions are asynchronous and can be executed in any order.
Thus, the user might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
Note that executing the bind action before the create will also fail but this case is prevented by the ioa::binding\_manager class.
Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.
A general solution is to observe the automaton and wait for it to be destroyed.

>>>>

\paragraph*{Declaring and initializing state variables.}
\fi
The state variables of an automaton are expressed as member variables of its corresponding class or class template.
\ifjournal
The four state variables of the asynch\_lcr\_automaton include the UID of the automaton, a queue of outgoing messages, and flags for reporting a change in status.
\begin{lstlisting}
private:
  const UID m_u;
  std::queue<UID> m_send;
  bool m_report;
  bool m_leader;
\end{lstlisting}
\fi
Member variables are declared private, which is in keeping with the notion that the state of each automaton is independent, and are initialized using a constructor.
The constructor can also bootstrap the scheduler using the ioa::schedule function discussed below.
\ifjournal
For the asynch\_lcr\_automaton, the flags are both set initially false and the UID is initialized and added to the send queue.
\begin{lstlisting}
public:
  asynch_lcr_automaton (const UID& u) :
    m_u (u),
    m_report (false),
    m_leader (false)
  {
    m_send.push (m_u);
    schedule ();
  }
\end{lstlisting}
The constructor also bootstraps the scheduler by scheduling all enabled local actions using the ioa::schedule function discussed below.
\fi

\paragraph*{Declaring/defining actions.}
Actions are member variables that model the \emph{action} 
\ifjournal
concept.
\else
concept\footnote{The term ``concept'' is used here in the sense of generic programming in C++, i.e., a collection of type requirements imposed by a C++ template.}.
\fi
Actions contain methods for preconditions, effects, and scheduling along with \emph{traits} that indicate the action's type (input, output, internal), value status, value type, parameter status, and parameter type.
These traits are used to guide the execution of the methods by the dispatcher and to supply values and parameters as necessary.
The traits are also used to check that input actions are bound to output actions with the same value status and type.
Bindings are checked for type compatibility at compile time using the generic programming facilities of C++.

@@

\paragraph*{Scheduling.}
Automata submit actions to the scheduler using the ioa::schedule function.
The ioa::schedule function can be called at any time.
However, automata are generally easier to read if scheduling calls are relegated to the scheduling functions.
A common idiom is for each scheduling function in an automaton to call a common member function that schedules all enabled local actions to avoid replicating scheduling logic.
The common scheduling function consists of calls to ioa::schedule guarded by preconditions as described in Section~\ref{scheduling}.

@@

\paragraph*{Dynamic composition.}
%% The asynch\_lcr\_automaton is a reusable component designed to be composed in the context of a larger system.
Dynamic composition is accomplished by creating child automata and bindings via ioa::automaton\_manager and ioa::binding\_manager objects.
\ifjournal
The tcp\_lcr\_automaton discussed in Section~\ref{case_study} creates a asynch\_lcr\_automaton and binds to it with the following code:
\begin{lstlisting}
ioa::automaton_manager<asynch_lcr_automaton<uuid> >*
lcr = ioa::make_automaton_manager (this,
  ioa::make_generator<asynch_lcr_automaton<uuid> > (
    m_u));
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::send,
  &m_self, &tcp_lcr_automaton::send_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::receive_lcr,
  lcr, &asynch_lcr_automaton<uuid>::receive);
ioa::make_binding_manager (this,
  lcr, &asynch_lcr_automaton<uuid>::leader,
  &m_self, &tcp_lcr_automaton::leader_lcr);
ioa::make_binding_manager (this,
  &m_self, &tcp_lcr_automaton::init_lcr,
  lcr, &asynch_lcr_automaton<uuid>::init);
\end{lstlisting}
\fi
%%Child automata are managed by ioa::automaton\_manager objects which can be created via the ioa::make\_automaton\_manager function.
An automaton manager object requires an ioa::automaton object for its system action implementation and a factory for allocating child automata.
%%Bindings are managed by ioa::binding\_manager objects which can be created via the ioa::make\_binding\_manager function.
A binding manager object requires an ioa::automaton object for its system action implementation, a manager object for the output automaton, a reference to the output action, a manger object for the input automaton, a reference to the input action, and parameters for the actions if required.
The ioa::automaton\_manager class and ioa::binding\_manager class also contain methods for destroying and unbinding respectively.
\ifjournal
The m\_self object of the example is an ioa::handle\_manager which can be used to bind to automata that already exist.
From the example, the tcp\_lcr\_automaton binds the asynch\_lcr\_automaton to itself.
The wary reader might notice that the ioa::automaton\_manager and ioa::binding\_managers in the example are dynamically allocated but never recorded and thus never freed.
Both classes are managed by the ioa::automaton and automatically deallocated when the automaton or binding they manage is destroyed or unbound respectively.
\fi

\paragraph*{Binding predicates.}
Recall that automata require the ability to inspect the current set of bindings to adhere to the I/O automata model.
The binding predicate supplied by ioa++ is the ioa::binding\_count function which returns the number of other actions bound to the specified action.
The value returned by ioa::binding\_count is zero for internal actions, zero or one for input actions, and non-negative for output actions.
Most often, ioa::binding\_count is used in the precondition of output actions to ensure that the value generated by the output action will be received if produced.
\ifjournal
This type of usage can be seen in the send action of the asynch\_lcr\_automaton:
\begin{lstlisting}
bool send_precondition () const {
  return !m_send.empty () &&
    ioa::binding_count (&asynch_lcr_automaton::send)
      != 0;
}
\end{lstlisting}
\fi

@@

\paragraph*{System action events.}
Determining the outcome of a system action event (e.g., that a child automaton was created) is necessary when there are constraints on the order in which system actions must occur.
For example, the Binding Manager mentioned in Section~\ref{system_action_section} waits until the automata that supply the output action and input action are created, before attempting to bind.
Similarly, an automaton might compose with an automaton that it does not manage, and therefore may expect the automaton to exist before attempting to bind.
To enable such behavior, user automata can \emph{observe} ioa::automaton\_manager objects and ioa::binding\_manager objects to receive events generated upon the completion of a system action.

\paragraph*{Error reporting.}
The event semantics of I/O automata provide a natural mechanism for reporting errors.
An automaton that encounters an error can use an output action to signal the error.
The automaton can continue processing if the error is transient or stop if the error is permanent.
In either case, the error is propagated to the automaton that is using the automaton that encountered an error.
Through this technique, errors can be propagated up the hierarchy and resolved at an appropriate level.
The two techniques for handling an error are either to reset or to replace the portion of the constellation that encountered the error.

\paragraph*{Handling errors by resetting.}
An automaton that can be reset has one or more output actions for reporting errors and one or more input actions for resetting.
Automaton designers should strive to write automata that are capable of being reset as this is the simplest error handling strategy.

\paragraph*{Handling errors by replacement.}
The non-deterministic and asynchronous nature of system actions complicates replacing an automaton.
To replace a child automaton, the parent automaton must destroy the child (thus dissolving all bindings with the child), create a new child, and bind to the new child.
Recall that system actions are asynchronous and can be executed in any order.
Thus, the user might request the actions in one order, e.g., destroy, create, bind, but the system may execute them in a different order, e.g., create, bind, destroy.
If the sequence executed by the system automaton is (create, bind, destroy) then the bind will fail because the user automaton's actions are still bound to the old child.
Note that executing the bind action before the create will also fail but this case is prevented by the ioa::binding\_manager class.
Thus, a user automaton must receive notification that the first child has been destroyed before (creating and) binding to the second child.
A general solution is to observe the automaton and wait for it to be destroyed.

### 2012-02-23 papers/paper2/related_work.tex
Removing material.

Leveraging a formal model of concurrency to make concurrent programming easier is a well established technique.
We consider three exemplars that are most relevant to the discussion in this paper: Communicating Sequential Processes, Actors, and I/O automata.
Of these, we selected the I/O automata model of Lynch and Tuttle\cite{lynch1987hierarchical} because of its support for features essential to developing distributed systems~\cite{lynch1996distributed}.
%% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.

The Communicating Sequential Processes (CSP) model~\cite{hoare1978communicating} is a popular technique for modeling concurrent systems.
The fundamental concurrency control primitive in CSP is the \emph{rendezvous} which allows processes to synchronize and possibly exchange values.
CSP was influential in the design of the occam programming language~\cite{jones1987programming} and continues to influence the design of modern programming languages, e.g., Go~\cite{go}.

An actor~\cite{agha1986actors} is a behavior that can perform a number of actions when receiving a message including (1) replacing itself with a new behavior, (2) sending messages to other actors, and (3) creating new actors.
The Actor model has influenced the design of languages such as Erlang~\cite{armstrong1996concurrent} and Scala~\cite{odersky2004overview}.

>>>>

Leveraging a formal model of concurrency to make concurrent programming easier is a well established technique.
We consider three exemplars that are most relevant to the discussion in this paper: Communicating Sequential Processes~\cite{hoare1978communicating}, Actors~\cite{agha1986actors}, and I/O automata~\cite{lynch1996distributed}.
Of these, we selected the I/O automata model because of its support for features essential to developing distributed systems.
%% Temporal logic proof techniques~\cite{manna1992temporal},~\cite{lamport1978time} also can be used to reason about I/O automata, towards developing \emph{correct} systems.
